---
title: "City Solvency Final Report - Draft"
author: "Cheikhou Kane"
date: "11/17/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Needed Libraries
library(tidyverse)
library(ggplot2)
library(wesanderson)
library(reshape2)

#Data
opeb.raw <- read.csv("OPEB_20.csv") 

pension.raw <- read_csv("PensionCalculation.csv")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Data Processing

opeb.raw$AAL_reported <- as.numeric(opeb.raw$AAL_reported)

##Adding Unfunded Liability (UAAL)
opeb.raw <- opeb.raw %>%
  mutate(UAAL_reported = AAL_reported - assets)

#Getting AAL with new discount rate
opeb.raw <- opeb.raw %>%
  mutate(AAL_standard =  (AAL_reported*((1+disc_rate)^15) )/ (1.0403)^15)

#UAAL with new discount rate
opeb.raw <- opeb.raw %>%
  mutate(UAAL_standard = AAL_standard - assets)

#corporate bond rate
CBR <- 0.0403
#standardized amoritization period
am_period <- 12

pension.raw <- pension.raw %>% 
            mutate(TPL = case_when(is.na(TotalPensionLiability) ~ 
                                     NetLiability + Assets),
                   TPL_minus1 = case_when(is.na(TPL_minus1) ~ 
                                            NL_minus1 + Assets), 
                   TPL_plus1 = case_when(is.na(TPL_plus1) ~ 
                                           NL_plus1 + Assets),
                   PercentOwed = PercentOwedTotal, 
                   Duration = (TPL_minus1 - TPL_plus1)/
                     (2*TPL*.01),
                   Convexity = (TPL_plus1 + TPL_minus1 - 
                                (2 * TPL))/ (2 * TPL * (0.01^2)),
                   Rate_Change = (CBR - DiscountRate),
                   Standardized_TPL =
                     ((am_period * Rate_Change) +
                     (0.5 * Convexity * Rate_Change^2 * 100) +
                       1)*TPL,
                   Calc_Type = 
                     case_when(!is.na(PerYear) ~ 
                              "Yearly Contrib", 
                              TRUE ~ "Percentage Contrib"),
                   Standardized_NL = case_when(!is.na(Assets) ~ Standardized_TPL - Assets,
                    is.na(Assets) ~ Standardized_TPL),
                   reported_pension = case_when(
                     Calc_Type == "Yearly Contrib" 
                     ~ round(Duration,0) * PerYear, 
                     (Calc_Type == "Percentage Contrib" & !is.na(Assets))
                     ~ NetLiability * PercentOwed, 
                   (Calc_Type == "Percentage Contrib" & is.na(Assets)) ~
                     TotalPensionLiability * PercentOwed),
                   standardized_pension = case_when(
                     Calc_Type == "Yearly Contrib" 
                     ~ am_period * PerYear, 
                     Calc_Type == "Percentage Contrib"
                     ~ Standardized_NL * PercentOwed))

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

#Summary data for OPEB
opeb.summary <- opeb.raw %>%
  group_by(city) %>%
  select(UAAL_reported,UAAL_standard, revenue, Pop, city_contrib) %>%
  summarise(UAAL_reported = sum(UAAL_reported),
            UAAL_standard = sum(UAAL_standard),
            Revenue = mean(revenue, na.rm = TRUE),
            Population = mean(Pop, na.rm = TRUE),
            city_contrib = sum(city_contrib, na.rm =TRUE))
```

```{r}
#Pension Summary

pension.summary <- pension.raw %>% 
                    group_by(City) %>% 
                    summarise(Pension_reported =
                                sum(reported_pension),
                              Pension_standard =
                                sum(standardized_pension))
```

```{r}
#combined dataframe with Pension + OPEB data
Pension.UL <- cbind(Pen_reported = pension.summary$Pension_reported, 
                    Pen_standard = pension.summary$Pension_standard)

total.summary <- cbind(opeb.summary, Pension.UL)

#Adding Total Lialbility
total.summary <- total.summary %>%
  mutate(total_reported = UAAL_reported + Pen_reported,
         total_standard = UAAL_standard + Pen_standard)

#Adding Per capita Calculation

total.summary$Revenue[!is.finite(total.summary$Revenue)] <- NA  #Clean out Revenue 
total.summary$Population[!is.finite(total.summary$Population)] <- NA #Clean out Population

total.summary <- total.summary %>%
  mutate(total_reported.capita = total_reported/Population,
         total_standard.capita = total_standard/Population)

total.summary <- total.summary %>%
  mutate(UAAL_reported.capita = UAAL_reported/Population,
         UAAL_standard.capita = UAAL_standard/Population)

total.summary <- total.summary %>%
  mutate(Pen_reported.capita = Pen_reported/Population,
         Pen_standard.capita = Pen_standard/Population)

#Adding share revenue

total.summary <- total.summary %>%
  mutate(OPEB.share = round(100*(city_contrib/Revenue),2))
```

# Introduction


## Overview of Liability Issues

## Discount Rates

## What is a city?


# I. Combined Liability: OPEB + Pension


## Total Liability

### Fig 1.a + 1.b OR Fig 1. 

```{r}
total.summary %>%
  select(city, total_reported, total_standard) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `total_reported`="Reported",
                           `total_standard`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value/1000), value/1000)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability (in millions)") + 
  ggtitle("Fig 1. Total Liability ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
```


## Total Per Capita

### Fig 2.a + 2.b OR Fig 2. 

```{r}
total.summary %>%
  select(city, total_reported.capita, total_standard.capita) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `total_reported.capita`="Reported",
                           `total_standard.capita`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value), value)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability per capita (in thousands)") + 
  ggtitle("Fig 2. Total Liability per capita ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
```


## Share of revenue

### Fig 3. 

**For this section I need the data on pension payments in 2017. The code to make the graph is ready however. ** 

# II. Pension Liability


## Total Pension Liability

### Fig 4.a + 4.b OR Fig 4. 

```{r}
total.summary %>%
  select(city, Pen_reported, Pen_standard) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `Pen_reported`="Reported",
                           `Pen_standard`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value/1000), value/1000)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability (in millions)") + 
  ggtitle("Fig 4. Pension Liability ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
  
```


## Total Pension Per Capita

### Fig 5.a + 5.b OR Fig 5. 

```{r}
total.summary %>%
  select(city, Pen_reported.capita, Pen_standard.capita) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `Pen_reported.capita`="Reported",
                           `Pen_standard.capita`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value), value)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability per capita (in thousands)") + 
  ggtitle("Fig 5. Pension Liability per capita ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
```


## Share of revenue

### Fig 6.

**For this section I need the data on pension payments in 2017. The code is ready however. ** 

# III. OPEB Liability


## Total OPEB Liability

### Fig 7.a + 7.b OR Fig 7. 

```{r}

total.summary %>%
  select(city, UAAL_reported, UAAL_standard) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `UAAL_reported`="Reported",
                           `UAAL_standard`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value/1000), value/1000)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability (in millions)") + 
  ggtitle("Fig 7. OPEB Liability ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
```


## Total OPEB Per Capita

### Fig 8.a + 8.b OR Fig 8. 

```{r}

total.summary %>%
  select(city, UAAL_reported.capita, UAAL_standard.capita) %>%
  melt(id.vars="city") %>%
  mutate(variable = recode(variable,
                           `UAAL_reported.capita`="Reported",
                           `UAAL_standard.capita`="Standardized")) %>%
  mutate(value = if_else(variable == "Reported", -(value), value)) %>%
  ggplot(aes(x = city, y = value, group = variable, fill = variable)) +
  geom_bar(stat = "identity", width = .9) +
  coord_flip() +
  #scale_x_discrete(limits = the_order) +
  scale_fill_manual(values=wes_palette(n=2, name="BottleRocket2")) +
  theme_classic() + xlab("City") + 
  ylab("Unfunded Liability per capita (in thousands)") + 
  ggtitle("Fig 8. OPEB Liability per capita ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Valuation Type")
```


## Share of revenue


### Fig 9.

```{r}
pal <- wes_palette("Zissou1", 10, type = "continuous")

total.summary %>%
  ggplot(aes(x = city, y = OPEB.share, fill = OPEB.share)) + geom_bar(stat = "identity") +
  scale_fill_gradientn(colours = pal) +
  theme_classic() + coord_flip() +
  xlab("City") + ylab("Share of Revenue") + 
  ggtitle("Fig 9. OPEB Benefit Payment as share of Revenue \n (2017 reported values)") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill = "Revenue Share (%)")
```

# Conclusion

